---
title: 'Step 1: Cleaning the Data'
author: "Al Xin"
date: "August 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

In this step, typos will be removed and the conditions for the experimental data will be standardized (298 K, water as a solvent, neutral pH, etc.) The cleaned data will be stored in the directory `data/clean`.

## Dependencies

There are several personal functions used throughout this project. They can be found and installed through GitHub using `devtools::install_github`.

```{r qsarr, message = F}
if (!require("devtools")) install.packages("devtools")
library(devtools)
install_github("awqx/qsarr")
```

Packages needed in this step include `data.table`, `dplyr`, and `stringr`.

```{r packages, message = F, warning = F}
packages <- c("data.table", "stringr", "tidyverse")
lapply(packages, library, character.only = T)
if(!dir.exists("data/clean")) dir.create("data/clean")
```

## Rekharsky and Inoue dataset

The data is read from the previous step. The data contains the following columns:
- `host`: α-, β-, or γ-cyclodextrin (written as `"alpha", "beta", "gamma"`) 
- `guest`: the ligand tested in the complex
- `dG`: the change in Gibbs free energy of the complexation, in kJ/mol
- `solvent`: a description of the solvent tested
- `temp`: the temperature of the experiment
- `ref`: the reference from the paper corresponding to the test

### Solvent conditions 

For consistency across observations, the data used will contain only observations conducted at 298 K in water as a solvent. Additionally, observations will be filtered to include observations from pH 6.8 to 7.5.

The original dataset contains 1367 observations. After cleaning for solvent conditions, the dataset contains 813 observations.

```{r ri_solvent}
# Reading the data from the previous step
ri_df <- readRDS("data/derived/ri-df.RDS")

# Removing unusual separators and spaces
ri <- ri_df %>%
  lapply(
    str_replace_all, pattern = "(\\â)|\u2212",
    replacement = "-") %>%
  lapply(str_replace_all, "(\\â+)|(Â·)|·|\u2009", " ") %>%
  lapply(str_replace_all, "\\s+", " ") %>%
  lapply(str_replace_all, "\\- ", "-") %>%
  as_tibble() %>%
  mutate(guest = str_replace_all(guest, "\\.", ","))

# Creating a column for pH
pH_regex <-  "[0-9]+\\.*[0-9]*"

# Splitting columns for solvent and solvent description
ri <- separate(
  ri,
  solvent, 
  c("solvent","solvent_desc"),
  sep = "(?=\\s*\\()", 
  extra = "merge", 
  fill = "right"
  )
ri <- ri %>% 
  mutate(
    pH = str_extract(
      solvent_desc, 
      "(pH\\s(\\<\\s)*[0-9]+(\\.[0-9]+)*)"),
    pH_range = str_extract(
      solvent_desc,
      "pH\\s[0-9]+\\.[0-9]+\\-[0-9]+(\\.[0-9]+)*")) %>%
  mutate(pH = ifelse(!is.na(pH_range), NA, pH)) %>%
  mutate(pH = as.numeric(str_extract(pH, pH_regex))) %>%
  separate(., pH_range, c("pH1", "pH2"), sep = "-") %>%
  mutate(
    pH1 = as.numeric(
      str_extract(pH1, pattern = pH_regex)),
    pH2 = as.numeric(
      str_extract(pH2, pattern = pH_regex))) %>%
  mutate(pH = ifelse(!is.na(pH1), (pH1 + pH2)/2, pH)) %>%
  mutate(pH = ifelse(is.na(pH), 7.0, pH)) %>%
  select(-pH1, -pH2) 


# Cleaning for temperature and solvent
ri <- ri %>%
  filter(
    solvent == "H2O", 
    temp == "298", 
    6.8 <= pH & pH < 7.5
  )
```

### Data reliability

According to the original Rekharsky and Inoue paper, datapoints with the footnote b, c, g, i, j, or m contain data that may be unreliable. These datapoints will be cleaned from the set. This leaves 785 observations.

```{r ri_ref}
ri <- filter(ri, !str_detect(ref, "b|c|g|i|j|m"))
```

### Removing salts

Hydrochloride salts will also be removed from the dataset. These are guest ligands that contain `HCl` in the name. This leaves 733 observations. 

```{r ri_salt}
ri <- filter(ri, !str_detect(guest, "\\sHCl"))
```

### Cyclodextrin hosts

This study will only examine α-, β-, or γ-cyclodextrin. Other varieties of cyclodextrin will be removed from the dataset. This leaves 638 observations.

```{r ri_cd}
ri <- filter(ri, str_detect(host, "(alpha)|(beta)|(gamma)"))
```

```{r ri_cd_cat, echo = F}
cat(
  "a-CD observations:", 
  sum(str_count(ri$host, "alpha")), 
  "\nb-CD:", 
  sum(str_count(ri$host, "beta")), 
  "\nc-CD:", 
  sum(str_count(ri$host, "gamma")))
```

### Correcting Gibbs free energy

Estimates of Gibbs free energy containing margins of error (e.g., plus or minus a measure) are cleaned to only contain the central estimate. 

```{r ri_dg}
ri <- ri %>%
  mutate(dG = str_remove(dG, "\\s\\±\\s.+")) %>%
  mutate(dG = as.numeric(dG))
```

### Saving data

The relevant columns for later steps are `host`, `guest`, and `dG`. These will be saved in `data/clean`.

```{r ri_save}
ri %>%
  select(host, guest, dG) %>%
  saveRDS("data/clean/ri.RDS")
```

