---
title: '2: Downloading ligand structures as SDFs'
author: "Al Xin"
date: "12/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The 3d structures of molecules can be downloaded as structure data files (SDFs) from the Chemical Identifier Resolver, a tool provided by the CADD group of the NCI. The URL is https://cactus.nci.nih.gov/chemical/structure

### Data organization

The downloaded SDFs will be in the directory `sdf`. Each data source will have its own subdirectory. All ligands will be compiled in the subdirectory `sdf/all`. Additionally, these subdirectories will be further divided by cyclodextrin type.

```{r dir}
cd_list <- c("alpha", "beta", "gamma")
df_list <- c("ri", "suzuki", "singh", "riss")
dir_list1 <- paste0("sdf/", df_list)
dir_list2 <- paste0(
  dir_list1, "/", 
  unlist(lapply(cd_list, rep, 4)))

if (!dir.exists("sdf")) {
  dir.create("sdf")
  lapply(dir_list1, dir.create)
  lapply(dir_list2, dir.create)
  dir.create("sdf/result")
}
```

Results from the downloads will be stored in `sdf/result`.

## Dependencies

There are several new functions used throughout this project. They can be found and installed through GitHub using `devtools::install_github`. Additionally, `dplyr` and `stringr` will be used here.

```{r dependencies, message = F}
library(devtools)
devtools::install_github("awqx/qsarr")
library(qsarr)
library(dplyr)
library(stringr)
```

### Timing

This is not necessary for the code to run, but to time the downloading, the library `tictoc` can be used.

```{r dependency_tictoc, eval = F}
install.packages("tictoc")
```

## Loading cleaned data

The cleaned data from the previous step is loaded to obtain the names of the ligands. 

```{r load_data, message = F}
ri <- readRDS("data/clean/ri.RDS")
suzuki <- readRDS("data/clean/suzuki.RDS")
singh <- readRDS("data/clean/singh.RDS")
riss <- readRDS("data/clean/riss.RDS")
```

## Rekharsky and Inoue

### SDF Download

The function `dwnld_mol_list` from `qsarr` will query NCI's database for molecules by name and download them into the indicated directory. The function output is a data frame indicating the status of the download. It is integrated in the function `dwnld_ligand` for better compatability with the data set.

Due to some output inconsistencies with  data frames, the function iterates using a `for` loop and not the `apply` family of functions. 

All 477 observations were succesfully downloaded. The download required 1198.11 seconds, or 19 minutes and 58.11 seconds.

```{r dwnld_ligand}
dwnld_ligand <- function(cd, p, df, ff) {
  dwnld_list <- list()
  if (length(cd) != length(p)) {
    message("Cyclodextrin list and list of directories are not equal")
    return(0)
  }
  for(i in 1:length(cd)) {
    guest <- filter(df, host == cd[i])$guest
    dwnld_list[[i]] <- dwnld_mol_list(
      mol_list = guest, 
      path = p[i], 
      file_format = ff
    ) %>% as_tibble()
  }
  dwnld_list <- Map(
    function(x, df) {
      mutate(df, host = x)
    }, 
    cd_list, 
    dwnld_list
  )
  do.call(rbind, dwnld_list)
}
```

```{r ri_dwnld, eval = F}
ri_dir_list <- dir_list2[str_detect(dir_list2, "ri/")]
if(require(tictoc)) tic()
ri_dwnld <- dwnld_ligand(cd_list, ri_dir_list, ri, "sdf")
if(require(tictoc)) toc()
saveRDS(ri_dwnld, "sdf/result/ri_dwnld.RDS")
```

```{r ri_dwnld_summary, echo = F}
ri_dwnld <- readRDS("sdf/result/ri_dwnld.RDS")
cat(
  "Successful downloads:", sum(ri_dwnld$downloaded), 
  "\nWarnings:", sum(ri_dwnld$warning), 
  "\nErrors:", sum(ri_dwnld$error)
)
```

## Suzuki

The Suzuki dataset only contains observations for α- and β- cyclodextrin. All 320 observations were successfully downloaded. The download required 250.58 seconds, or 4 minutes and 10.58 seconds. 

```{r suzuki_dwnld, eval = F}
suzuki_dir_list <- dir_list2[str_detect(dir_list2, "suzuki")]
if(require(tictoc)) tic()
suzuki_dwnld <- dwnld_ligand(cd_list, suzuki_dir_list, suzuki, "sdf")
if(require(tictoc)) toc()
saveRDS(suzuki_dwnld, "sdf/result/suzuki_dwnld.RDS")
```

```{r suzuki_dwnld_summary, echo = F}
suzuki_dwnld <- readRDS("sdf/result/suzuki_dwnld.RDS")
cat(
  "Successful downloads:", sum(suzuki_dwnld$downloaded), "\n",
  "Warnings:", sum(suzuki_dwnld$warning), "\n",
  "Errors:", sum(suzuki_dwnld$error)
)
```

## Singh

The Singh data only contains observations for γ-cyclodextrin. The names of the guest molecules need to be all lowercase to be properly queried. All 112 observations were successfully downloaded. The download required 105.31 seconds, or 1 minute and 45.31 seconds. 

```{r singh_dwnld, eval = F}
singh <- mutate(singh, guest = tolower(guest))
singh_dir_list <- dir_list2[str_detect(dir_list2, "singh")]
if(require(tictoc)) tic()
singh_dwnld <- dwnld_ligand(cd_list, singh_dir_list, singh, "sdf")
if(require(tictoc)) toc()
saveRDS(singh_dwnld, "sdf/result/singh_dwnld.RDS")
```

```{r singh_dwnld_summary, echo = F}
singh_dwnld <- readRDS("sdf/result/singh_dwnld.RDS")
cat(
  "Successful downloads:", sum(singh_dwnld$downloaded), "\n",
  "Warnings:", sum(singh_dwnld$warning), "\n",
  "Errors:", sum(singh_dwnld$error)
)
```

## Compiled data

The compiled data (RISS) contains observations for all three cyclodextrin types. All 766 observations were successfully downloaded. The download took 610.64 seconds, or 10 minutes and 10.64 seconds.

```{r riss_dwnld, eval = F}
riss_dir_list <- dir_list2[str_detect(dir_list2, "riss/")]
if(require(tictoc)) tic()
riss_dwnld <- dwnld_ligand(cd_list, riss_dir_list, riss, "sdf")
if(require(tictoc)) toc()
saveRDS(riss_dwnld, "sdf/result/riss_dwnld.RDS")
```

```{r riss_dwnld_summary, echo = F}
riss_dwnld <- readRDS("sdf/result/riss_dwnld.RDS")
cat(
  "Successful downloads:", sum(riss_dwnld$downloaded), "\n",
  "Warnings:", sum(riss_dwnld$warning), "\n",
  "Errors:", sum(riss_dwnld$error)
)
```