---
title: '02. Downloading ligand structures as SDFs'
author: "Al Xin"
date: "12/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, eval = F)
```

## Overview

The 3d structures of molecules can be downloaded as structure data files (SDFs) from the Chemical Identifier Resolver, a tool provided by the CADD group of the NCI. The URL is https://cactus.nci.nih.gov/chemical/structure

### Data organization

The downloaded SDFs will be in the directory `sdf`. Each data source will have its own subdirectory. All ligands will be compiled in the subdirectory `sdf/all`. Additionally, these subdirectories will be further divided by cyclodextrin type.

```{r dir, eval = T}
cd_list <- c("alpha", "beta", "gamma")
df_list <- c("ri", "suzuki", "singh", "riss")
dir_list1 <- paste0("sdf/", df_list)
dir_list2 <- paste0(
  dir_list1, "/", 
  unlist(lapply(cd_list, rep, 4)))

if (!dir.exists("sdf")) {
  dir_create("sdf")
  lapply(dir_list1, dir_create)
  dir_create("sdf/result")
}

if (!dir.exists(dir_list2[1])) {
  lapply(dir_list2, dir_create)
}
```

Results from the downloads will be stored in `sdf/result`.

## Dependencies

There are several new functions used throughout this project. They can be found and installed through GitHub using `devtools::install_github`. Additionally, `dplyr` and `stringr` will be used here.

```{r dependencies, message = F, eval = T}
if (!require("qsarr")) {
 devtools::install_github(
  "awqx/qsarr", 
  upgrade = "never", 
  quiet = T
  ) 
}

library(qsarr)
library(dplyr)
library(stringr)
```

### Timing

This is not necessary for the code to run, but to time the downloading, the library `tictoc` can be used.

```{r dependency_tictoc}
library(tictoc)
```

## Loading cleaned data

The cleaned data from the previous step is loaded to obtain the names of the ligands. 

```{r load_data, message = F}
ri <- readRDS("affinity/clean/ri.RDS")
suzuki <- readRDS("affinity/clean/suzuki.RDS")
singh <- readRDS("affinity/clean/singh.RDS")
riss <- readRDS("affinity/clean/riss.RDS")
```

## Rekharsky and Inoue

### SDF Download

The function `dwnld_mol_list` from `qsarr` will query NCI's database for molecules by name and download them into the indicated directory. The function output is a data frame indicating the status of the download. It is integrated in the function `dwnld_ligand` for better compatability with the data set.

Due to some output inconsistencies with  data frames, the function iterates using a `for` loop and not the `apply` family of functions. 

All 477 observations were succesfully downloaded. The download required 1198.11 seconds, or 19 minutes and 58.11 seconds.

```{r dwnld_ligand}
dwnld_ligand <- function(cd, p, df, ff) {
  dwnld_list <- list()
  if (length(cd) != length(p)) {
    message("Cyclodextrin list and list of directories are not equal")
    return(0)
  }
  for(i in 1:length(cd)) {
    guest <- filter(df, host == cd[i])$guest
    dwnld_list[[i]] <- dwnld_mol_list(
      mol_list = guest, 
      path = p[i], 
      file_format = ff
    ) %>% as_tibble()
  }
  dwnld_list <- Map(
    function(x, df) {
      mutate(df, host = x)
    }, 
    cd_list, 
    dwnld_list
  )
  do.call(rbind, dwnld_list)
}
```

```{r ri_dwnld, eval = F}
ri_dir_list <- dir_list2[str_detect(dir_list2, "ri/")]
if(require(tictoc)) tic()
ri_dwnld <- dwnld_ligand(cd_list, ri_dir_list, ri, "sdf")
if(require(tictoc)) toc()
saveRDS(ri_dwnld, "sdf/result/ri_dwnld.RDS")
```

```{r ri_dwnld_summary, echo = F, eval = T}
ri_dwnld <- readRDS("sdf/result/ri_dwnld.RDS")
cat(
  "Successful downloads:", sum(ri_dwnld$downloaded), 
  "\nWarnings:", sum(ri_dwnld$warning), 
  "\nErrors:", sum(ri_dwnld$error)
)
```

## Suzuki

The Suzuki dataset only contains observations for α- and β- cyclodextrin. All 320 observations were successfully downloaded. The download required 250.58 seconds, or 4 minutes and 10.58 seconds. 

```{r suzuki_dwnld, eval = F}
suzuki_dir_list <- dir_list2[str_detect(dir_list2, "suzuki")]
if(require(tictoc)) tic()
suzuki_dwnld <- dwnld_ligand(cd_list, suzuki_dir_list, suzuki, "sdf")
if(require(tictoc)) toc()
saveRDS(suzuki_dwnld, "sdf/result/suzuki_dwnld.RDS")
```

```{r suzuki_dwnld_summary, echo = F, eval = T}
suzuki_dwnld <- readRDS("sdf/result/suzuki_dwnld.RDS")
cat(
  "Successful downloads:", sum(suzuki_dwnld$downloaded), "\n",
  "Warnings:", sum(suzuki_dwnld$warning), "\n",
  "Errors:", sum(suzuki_dwnld$error)
)
```

## Singh

The Singh data only contains observations for γ-cyclodextrin. The names of the guest molecules need to be all lowercase to be properly queried. All 112 observations were successfully downloaded. The download required 105.31 seconds, or 1 minute and 45.31 seconds. 

```{r singh_dwnld, eval = F}
singh <- mutate(singh, guest = tolower(guest))
singh_dir_list <- dir_list2[str_detect(dir_list2, "singh")]
if(require(tictoc)) tic()
singh_dwnld <- dwnld_ligand(cd_list, singh_dir_list, singh, "sdf")
if(require(tictoc)) toc()
saveRDS(singh_dwnld, "sdf/result/singh_dwnld.RDS")
```

```{r singh_dwnld_summary, echo = F, eval = T}
singh_dwnld <- readRDS("sdf/result/singh_dwnld.RDS")
cat(
  "Successful downloads:", sum(singh_dwnld$downloaded), "\n",
  "Warnings:", sum(singh_dwnld$warning), "\n",
  "Errors:", sum(singh_dwnld$error)
)
```

## Compiled data

The compiled data (RISS) contains observations for all three cyclodextrin types. All 766 observations were successfully downloaded. The download took 610.64 seconds, or 10 minutes and 10.64 seconds.

```{r riss_dwnld}
riss_dir_list <- dir_list2[str_detect(dir_list2, "riss/")]
if(require(tictoc)) tic()
riss_dwnld <- dwnld_ligand(cd_list, riss_dir_list, riss, "sdf")
if(require(tictoc)) toc()
saveRDS(riss_dwnld, "sdf/result/riss_dwnld.RDS")
```

```{r riss_dwnld_summary, echo = F, eval = T}
riss_dwnld <- readRDS("sdf/result/riss_dwnld.RDS")
cat(
  "Successful downloads:", sum(riss_dwnld$downloaded), "\n",
  "Warnings:", sum(riss_dwnld$warning), "\n",
  "Errors:", sum(riss_dwnld$error)
)
```

## Combining SDFs

For certain applications, like the `rcdk` package, it is convenient to have SDFs compiled into a single SDF. The functions `qsarr::combine_sdf` and `qsarr::write_sdf` can accomplish this. These files will be saved in `sdf/compiled`.

```{r combine_sdf, eval = T}
if (!dir.exists("sdf/combined")) dir_create("sdf/combined")

# Recall that dir_list2 holds the directories for where
# all the SDFs are saved
sdf_list <- lapply(
  dir_list2, 
  combine_sdf
)

# Retain appropriate file names and convert them into files
sdf_file <- dir_list2[which(is.na(as.logical(sdf_list)))] %>%
  str_remove("sdf/") %>%
  str_replace("/", "_") %>%
  paste0("sdf/combined/", .)

# Remove zero-length elements of the list
sdf_list <- sdf_list[which(is.na(as.logical(sdf_list)))]

# Save the SDF files
mapply(
  write_sdf, 
  sdf = sdf_list, 
  path = sdf_file
)
```

### Removing names

Certain names in the SDFs may be problematic later due to non-ASCII characters or commas. A separate directory of molecules without names will be created to address this (`sdf/no_name`). The names will be stored as an RDS file in this directory. 

```{r combine_noname}
if (!dir.exists("sdf/no_name")) dir_create("sdf/no_name")

# Recall that dir_list2 holds the directories for where
# all the SDFs are saved
sdf_noname <- lapply(
  dir_list2, 
  combine_sdf, 
  rm_name = T
)

# Retain appropriate file names and convert them into files
sdf_file <- dir_list2[which(is.na(as.logical(sdf_noname)))] %>%
  str_remove("sdf/") %>%
  str_replace("/", "_") %>%
  paste0("sdf/no_name/", .)

# Remove zero-length elements of the list
sdf_noname <- sdf_noname[which(is.na(as.logical(sdf_noname)))]

# Save the SDF files
mapply(
  write_sdf, 
  sdf = sdf_noname, 
  path = sdf_file
)
```

