---
title: '03. Calculating chemical descriptors using CDK for R'
author: "Al Xin"
date: "12/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

## Overview

The package `rcdk` links the Java library CDK (Chemistry Development Kit) to R. The CDK will be used for generating chemical descriptors for QSAR building.

### Data organization 

Descriptors will be saved in the directory `desc`. File names will be symmetric with file names for combined SDFs, 

```{r desc_dir, message = F, eval = T}
if (!dir.exists("desc")) dir_create("desc")
```

## Dependencies

The package `rcdk` is required here. This will automatically load `rJava` as well. 

```{r pkg, message = F, eval = T}
library(rcdk)
```

## Loading molecules

Recall that the combiend SDFs are in the directory `sdf/combined`. Using `rcdk::load.molecules` we can create `jobjRef` objects (provided by `rJava`) for use with CDK.

We will then subset the elements of the list according to the data source. 

```{r load_mol}
sdf_file <- list.files("sdf/combined", full.names = T)
sdf_name <- list.files("sdf/combined") %>%
  str_remove("\\.(?i)sdf$")

# Creating a list of objects with rJava
# Names correspond to the combined SDF names
mol_list <- lapply(
  sdf_file, 
  load.molecules
)
names(mol_list) <- sdf_name
```

## Chemical descriptors

`rcdk` provides a list of all descriptor categories through `get.desc.categories()`. There is no single category for all descriptors. To evaluate all available descriptors, we have to unlist and combine the unique descriptors from the available categories. 

Evaluation of descriptors will irreversibly fail with BCUT descriptors, which correspond to eigenvalues of the adjacency matrix of the molecules, described at https://vcclab.org/lab/indexhlp/bcutdes.html. As a result, BCUT descriptors, corresponding to index 2 in the descriptor list, are removed from evaluation. 

```{r all_desc}
all_desc <- unique(
  unlist(
    sapply(
      get.desc.categories(), 
      get.desc.names
    )
  )
)
all_desc <- all_desc[-2]
```

We can then evaluate the descriptors using `eval.desc`. If a descriptor cannot be calculated, a `NA` or `NaN` will be returned.

The descriptors are returned as a data frame of descriptors. This process removes the names of the molecules and they have to be replaced using `get.title`. 

```{r eval_desc}
desc <- lapply(
  mol_list, 
  eval.desc, 
  which.desc = all_desc
)

mol_name <- lapply(mol_list, function(x) sapply(x, get.title))

desc_df <- Map(
  function(x, y) mutate(x, guest = y), 
  x = desc, 
  y = mol_name
)
```

We can then save the data frames of descriptors in `desc/cdk`. The descriptors are saved in a subfolder because we may incorporate other descriptor sources later (e.g., PaDEL-descriptor) 

```{r save_desc}
if (!dir.exists("desc/cdk")) dir_create("desc/cdk")
desc_name <- paste0("desc/cdk/", names(desc_df), ".RDS")
save_result <- mapply(
  function(x, y) saveRDS(x, y), 
  x = desc_df, 
  y = desc_name
)
```
