---
title: "05. Creating external validation sets"
author: "Al Xin"
date: "12/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

## Overview

According to QSAR best practices [^1], models should be tested on external validation sets after training. This is distinct from a "test" set in the model-building process, which can be created from a split in the data or *k*-fold cross-validation, for example. Test sets are used to select models and may provide an inflated accuracy if interpreted as-is. 

For each file of descriptors, 15% of the data will be set aside for external validation. External validation sets will be saved in the directory `extval/`. The sets used in model evaluation will be in the directory `trn/`. The directory structure of both of these directories will mirror that of `desc/` or `join/`. 

```{r dir_create}
if (!dir.exists("extval")) dir_create("extval")
if (!dir.exists("trn")) dir_create("trn")

# Create appropriate directories
trn_dir <- paste0("trn/", list.files("desc", include.dirs = T))
extval_dir <- paste0("extval/", list.files("desc", include.dirs = T))

if (!dir.exists(trn_dir[1])) {
  lapply(trn_dir, dir_create)
  lapply(extval_dir, dir_create)
}
```

## Dependencies

```{r dependencies}
library(qsarr)
library(stringr)
library(dplyr)
```

## Reading data

The input data will be the data obtained from joining experimental affinity results and the chemical descriptors. These files have been saved in the directory `join/`. The functions `read_desc` and `read_desc_list` from `qsarr` can handle the reading of files.

```{r}
join_file <- list.files(
  "join", 
  recursive = T, 
  full.names = T
)

join_list <- read_desc_list(join_file, quiet = T)
```

## Splitting data

The following uses `qsarr::make_extval`, which returns a list of data frames containing an entry for the retained data (named `"trn"`) and an entry for external validation data (named `"extval"`.) The data frames are then saved in the appropriate directories. 

```{r}
# Create a list, where each entry is a list with two elements
  # trn: The data retained for training
  # extval: The data for external validation
extval_list <- lapply(
  join_list, 
  make_extval, 
  p = 0.15, 
  seed = 20220101
)

# Generate appropriate file names
extval_file <- str_replace(join_file, "^join/", "extval/")
trn_file <- str_replace(join_file, "^join/", "trn/")

# Assigning to variable to prevent output
ev_dwnld <- Map(
  function(df, t_dir, e_dir) {
    saveRDS(df$trn, t_dir)
    saveRDS(df$extval, e_dir)
  }, 
  df = extval_list, 
  t_dir = trn_file, 
  e_dir = extval_file
)
```

[^1]: Alexander Tropsha (2010). Best Practices for QSAR Model Development, Validation, and Exploitation, *Molecular Informatics*. https://doi.org/10.1002/minf.201000061

