---
title: "08. Random forest proof of concept"
author: "Al Xin"
date: "1/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

## Overview

Models built on all predictor variables will be saved in `model/rf/all_var/`. 
Models trained on variables selected from RFE in 09 will be saved in `model/rf/rfe_var`. 

```{r directories}
desc_src <- c("cdk", "mordred", "ochem", "padel")

if (!dir.exists("model")) dir.create("model")
if (!dir.exists("model/rf")) {
  dir.create("model/rf")
  dir.create("model/rf/all_var")
  dir.create("model/rf/rfe_var")
  dir_create_result <- sapply(paste0("model/rf/all_var/", desc_src), dir.create)  
  dir_create_result <- sapply(paste0("model/rf/rfe_var/", desc_src), dir.create)  
}
```

### General outline for model building

## Dependencies

```{r package}
# library(caret)
library(dplyr)
# # library(stringr)
# library(tibble)
# library(tidyr)
library(qsarr)
library(stringr)
library(tidyverse)
# library(randomForest)
library(parallel)
library(doParallel)

theme_set(theme_bw())

```

## All variables

For creating models with all the variables, the descriptors are read as separate 
lists. This makes it clearer that different possible tuning parameters are used
based on the number of descriptors available. 

Previously in step 08 (random forest proof of concept), tuning yielded that 400
trees was sufficient for training. For a margin of error, 500 trees will be used.

```{r cdk_all}
# Read preprocessed descriptors
# Remove guest names for training
cdk_list <- read_desc_list(
  list.files("preprocess/cdk", full.names = T), 
  quiet = T
)

padel_list <- read_desc_list(
  list.files("preprocess/padel", full.names = T), 
  quiet = T
)

mordred_list <- read_desc_list(
  list.files("preprocess/mordred", full.names = T), 
  quiet = T
)

ochem_list <- read_desc_list(
  list.files("preprocess/ochem", full.names = T), 
  quiet = T
) 

data_name <- names(cdk_list)


# Parallel processing setup
num_core <- max(1, detectCores() - 1)
cl <- makeCluster(num_core)
registerDoParallel(cl)
cl_lib <- clusterEvalQ(cl, library("qsarr"))
# cl_lib <- clusterEvalQ(cl, library("tibble"))
# cl_lib <- clusterEvalQ(cl, library("randomForest"))
clusterSetRNGStream(cl = cl, 20210116)

# library(caret)
# library(dplyr)
# library(tibble)
# library(tidyr)
# library(qsarr)
# library(randomForest)
# library(parallel)
# library(doParallel)
# clusterExport(
#   cl = cl, 
#   varlist = c(
#     "tune", 
#     "tune.randomForest", 
#     "tune_helper",
#     "best_summary_stat", 
#     "eval_model", 
#     "eval_model_rep", 
#     "normalize_"
#   )
# )
cdk_all <- parLapply(
  cl = cl,
  cdk_list, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(4, 16, 32, 64), 
  nodesize = c(2, 4, 8)
)

cdk_all_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(cdk_all[[x]], paste0("model/rf/all_var/cdk/", x, ".RDS"))
  }
)

padel_all <- parLapply(
  cl = cl,
  padel_list, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(4, 20, 40, 100), 
  nodesize = c(2, 4, 8)
)

padel_all_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(padel_all[[x]], paste0("model/rf/all_var/padel/", x, ".RDS"))
  }
)

mordred_all <- parLapply(
  cl = cl,
  mordred_list, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(4, 20, 40, 100), 
  nodesize = c(2, 4, 8)
)

mordred_all_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(mordred_all[[x]], paste0("model/rf/all_var/mordred/", x, ".RDS"))
  }
)

ochem_all <- parLapply(
  cl = cl,
  ochem_list, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(4, 20, 60, 150), 
  nodesize = c(2, 4, 8)
)

ochem_all_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(ochem_all[[x]], paste0("model/rf/all_var/ochem/", x, ".RDS"))
  }
)

stopCluster(cl = cl)
```
```{r rfe}
# Read preprocessed descriptors
# Remove guest names for training
cdk_list_rfe <- read_desc_list(
  list.files("rfe/df/cdk", full.names = T), 
  quiet = T
)

padel_list_rfe <- read_desc_list(
  list.files("rfe/df/padel", full.names = T), 
  quiet = T
)

mordred_list_rfe <- read_desc_list(
  list.files("rfe/df/mordred", full.names = T), 
  quiet = T
)

ochem_list_rfe <- read_desc_list(
  list.files("rfe/df/ochem", full.names = T), 
  quiet = T
) 

data_name <- names(cdk_list_rfe)

set.seed(20210114)
cdk_rfe <- lapply(
  cdk_list_rfe, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500,
  replace = c(T, F),
  mtry = c(2, 4, 8, 12),
  nodesize = c(2, 4, 8)
)

cdk_rfe_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(cdk_rfe[[x]], paste0("model/rf/rfe_var/cdk/", x, ".RDS"))
  }
)


padel_rfe <- lapply(
  padel_list_rfe, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(2, 4, 8, 12), 
  nodesize = c(2, 4, 8)
)

padel_rfe_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(padel_rfe[[x]], paste0("model/rf/rfe_var/padel/", x, ".RDS"))
  }
)


mordred_rfe <- lapply(
  mordred_list_rfe, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(2, 4, 8, 12), 
  nodesize = c(2, 4, 8)
)

mordred_rfe_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(mordred_rfe[[x]], paste0("model/rf/rfe_var/mordred/", x, ".RDS"))
  }
)


ochem_rfe <- lapply(
  ochem_list_rfe, 
  tune, 
  method = "rf", 
  resp = "dG",
  ignore_col = "guest",
  nfold = 10,
  nrep = 10,
  ntree = 500, 
  replace = c(T, F),
  mtry = c(2, 4, 8, 12), 
  nodesize = c(2, 4, 8)
)

ochem_rfe_save <- lapply(
  data_name, 
  function(x) {
    saveRDS(ochem_rfe[[x]], paste0("model/rf/rfe_var/ochem/", x, ".RDS"))
  }
)
```

## External validation

```{r}
desc_name <- c("cdk", "mordred", "ochem", "padel")
data_name <- str_remove(list.files("desc/cdk"), "\\.RDS$")
```
```{r}
ev_list <- lapply(
  paste0("extval_pp/", desc_name), 
  function(x) {
    lapply(
      list.files(x, full.names = T), 
      readRDS
    ) %>%
      setNames(nm = data_name)
  }
) %>%
  setNames(nm = desc_name)
```
```{r}
rf_allvar <- lapply(
  paste0("model/rf/all_var/", desc_name), 
  function(x) {
    lapply(
      list.files(x, full.names = T), 
      readRDS
    ) %>%
      setNames(nm = data_name)
  }
) %>%
  setNames(nm = desc_name)

ev_allvar <- lapply(
  desc_name, 
  function(x) {
    tune_sublist <- rf_allvar[[x]]
    ev_sublist <- ev_list[[x]]
    lapply(
      data_name, 
      function(y) {
        pred <- predict(tune_sublist[[y]], ev_sublist[[y]])
        data.frame(
          guest = ev_sublist[[y]]$dG, 
          in_ad = ev_sublist[[y]]$in_ad,
          obs = ev_sublist[[y]]$dG, 
          pred = pred
        )
      }
    ) %>%
      setNames(data_name)
  }
) %>% 
  setNames(desc_name)

# All results, regardless of applicability domain
ev_allvar_result <- lapply(
  ev_allvar, 
  function(x) {
    lapply(
      x, 
      defaultSummary
    ) %>%
      setNames(data_name)
  }
) %>%
  setNames(desc_name)

# Accuracy only on applicability domain
ev_allvar_result_ad <- lapply(
  ev_allvar, 
  function(x) {
    lapply(
      x, 
      function(y) {
        y <- y[y$in_ad, ]
        defaultSummary(y)
      }
    ) %>%
      setNames(data_name)
  }
) %>% 
  setNames(desc_name)

ev_allvar_df <- data.table::rbindlist(
  lapply(
    unlist(ev_allvar_result, recursive = F), 
    function(x) {
      data.frame(as.list(x))
    }
  )
) %>%
  data.frame(
    desc_src = unlist(lapply(desc_name, rep, length(data_name))), 
    data_src = rep(data_name, length(desc_name))
  ) %>%
  pivot_longer(RMSE:MAE, names_to = "summary_stat")

ev_allvar_ad_df <- data.table::rbindlist(
  lapply(
    unlist(ev_allvar_result_ad, recursive = F), 
    function(x) {
      data.frame(as.list(x))
    }
  )
) %>%
  data.frame(
    desc_src = unlist(lapply(desc_name, rep, length(data_name))), 
    data_src = rep(data_name, length(desc_name))
  ) %>%
  pivot_longer(RMSE:MAE, names_to = "summary_stat")

ggplot(ev_allvar_df, aes(x = data_src, y = value, fill = desc_src)) + 
  geom_col(position = "dodge") + 
  facet_grid(summary_stat ~ ., scales = "free")

ggplot(ev_allvar_ad_df, aes(x = data_src, y = value, fill = desc_src)) + 
  geom_col(position = "dodge") + 
  facet_grid(summary_stat ~ ., scales = "free")
# rf_tst <- rf_allvar[[1]]
# ev_tst <- ev_list[[1]]
# pred_tst <- lapply(
#   names(rf_tst), 
#   function(x) {
#     # rf_model <- rf_tst[[x]]$model
#     # predict(rf_model, ev_tst[[x]])
#     predict(rf_tst[[x]], ev_tst[[x]])
#   }
# )
```

```{r}
# print.tune <- function(tune_obj) {
#   cat(
#     "Model type: ", class(tune_obj$model), "\n\n",
#     "Tuned parameters: ", "\n",
#     sapply(
#       names(tune_obj$param),
#       function(x) {
#         paste0("\t", x, ": ", tune_obj$param[[x]], "\n")
#       }
#     ), "\n",
#     "Number of predictors: ", length(tune_obj$pred_name), "\n",
#     "Predictors (first 10): ",
#     tune_obj$pred_name[1:(min(10, length(tune_obj$pred_name)))],
#     "\n", "\n",
#     "Model performance: ", "\n"
#   )
#   print(tune_obj$result_summary)
# }
# 
# predict.tune <- function(tune_obj, ...) {
#   # if (length(class(tune_obj$model)) > 1) {
#   #   class(tune_obj$model) <-
#   #     class(tune_obj$model)[length(class(tune_obj$model))]
#   # }
#   # best_model <- tune_obj$model
#   # UseMethod("predict", best_model)
#   predict(tune_obj$model, ...)
# }
```

